Uncovering Ferry Traffic Patterns: A Data-Driven Study of Toronto Island Ticket Sales and Redemptions

### 2024-09-27 outputs/paper/paper.qmd
made a few amendments on the main text
add a few details on data analsis
add more informations on the data cleaning part in appendix section

The dataset used in this analysis is the 2024 installment of “Toronto Island Ferry Ticket Counts” obtained from OpenDataToronto (@varga1987toronto). This dataset contains information about the number of ferry tickets redeemed and sold for the Toronto Island Ferry system, which serves as a critical transportation link between downtown Toronto and the Toronto Islands.

@@

The data was processed and analyzed using the R programming language (@citeR). The tidyverse (@tidyverse2019) and opendatatoronto packages were employed to download, clean, and prepare the dataset for analysis. The resulting cleaned dataset was then utilized to generate the visualizations and statistical analyses presented in this paper using a collection of libraries which included the following packages:

- ggplot2 [@wickham2016ggplot2]
- knitr [@xie2023knitr]
- readr [@wickham2023readr]

@@

The dataset used in this analysis provides timestamped records of ticket redemptions and sales for the Toronto Island Ferry system, allowing for a detailed time series analysis of ferry usage patterns. As illustrated in @fig-linePlot1, ferry usage fluctuates significantly throughout the year, with peak activity consistently observed between July and September. This pattern corresponds to the increased passenger demand during the summer holiday season and major events. In 2020, ticket demand saw a sharp decline, likely due to the impact of the COVID-19 pandemic(@urbanyi2020maritime), but it showed signs of gradual recovery starting in 2021 as restrictions eased and tourism began to rebound in Toronto(@travel_gov_2024). By identifying these peak traffic periods in advance, ferry operators can implement targeted operational strategies, such as scheduling additional ferry trips or introducing crowd control measures to minimize the risk of overcrowding or service disruptions. Identifing these high-demand periods can ensure smoother ferry operations and enhance the overall passenger experience.

```{r}
#| label: fig-linePlot1
#| fig-cap: Redemption and Sales Counts Over Time in days
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)

@@

@fig-linePlot2 below shows the Redemption and Sales Counts in months over years 加一段描述plot2

```{r}
#| label: fig-linePlot2
#| fig-cap: Redemption and Sales Counts Over Time in months
#| warnings: false
#| message: false
#| echo: false
#| fig-pos: "H"  #fix the plot in current place when render
#| fig-width: 8
#| fig-height: 4
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)

@@

@fig-linePlot3 on the next page compares the sales count and redemption count in year 2021 and 2023

1 emm这里写一点描述fig3的对比？，render后看页面排版

```{r}
#| label: fig-linePlot3
#| fig-cap: Redemption and Sales Counts Over Time in 2021 and 2023
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"  #fix the plot in current place when render
#| fig-width: 8
#| fig-height: 4

@@

# Plot for 2023
plot_2023 <- ggplot(data_2023, aes(x = Timestamp)) +
  geom_line(aes(y = Redemption.Count), color = "#A977A6", linewidth = 1.2) +
  geom_line(aes(y = -Sales.Count), color = "#C1A9D2", linewidth = 1.2) +
  scale_y_continuous(limits = c(-y_max, y_max), sec.axis = sec_axis(~ -., name = "Sales Count")) +
  scale_x_datetime(date_breaks = "4 months", date_labels = "%b") +  # Show only 3 labels
  labs(y = "Redemption Count", x = "Month", title = "2023") +
  theme_minimal()

# Plot for 2021
plot_2021 <- ggplot(data_2021, aes(x = Timestamp)) +
  geom_line(aes(y = Redemption.Count), color = "#F49600", linewidth = 1.2) +
  geom_line(aes(y = -Sales.Count), color = "#F9CB80", linewidth = 1.2) +
  scale_y_continuous(limits = c(-y_max, y_max), sec.axis = sec_axis(~ -., name = "Sales Count")) +
  scale_x_datetime(date_breaks = "4 months", date_labels = "%b") +  # Show only 3 labels
  labs(y = "Redemption Count", x = "Month", title = "2021") +
  theme_minimal()

# Arrange the two plots side by side
grid.arrange(plot_2021, plot_2023, ncol = 2)

```
As shown in @fig-linePlot4, the x-axis denotes the timestamps of each record, while the y-axis represents the number of tickets processed. The blue line corresponds to the redemption counts, and the green line represents the sales counts, plotted inversely to emphasize the contrast between the two trends. The symmetrical y-axis allows for a clearer comparison between peaks and troughs in both counts over time.

```{r}
#| echo: false 
#| warning: false
#| message: false
#| label: fig-linePlot4
#| fig-cap: Redemption and Sales Counts Over Time in years
#| fig-pos: "H"  #fix the plot in current place when render
#| fig-width: 6
#| fig-height: 2.5

@@

# Create the plot
ggplot(data, aes(x = Timestamp)) +
  # Line plot for Redemption Count
  geom_line(aes(y = Redemption.Count), color = "#F7C4C1", linewidth = 1.2) +
  # Line plot for Sales Count (inverted)
  geom_line(aes(y = -Sales.Count), color = "#83A0BE", linewidth = 1.2) +

  # Set y-axis limits symmetrically and apply the secondary axis
  scale_y_continuous(limits = c(-y_max, y_max),
                     breaks = seq(-y_max, y_max, by = 1000),
                     sec.axis = sec_axis(~ -., 
                                         name = "Sales Count",
                                         breaks = seq(y_max, -y_max, by = -1000))) +

@@

  # Apply minimal theme and customize axis titles
  theme_minimal() +
  theme(
    # Color the right-side axis title (Sales Count) to match the orange line
    axis.title.y.right = element_text(color = "#F7C4C1", size = 12),
    # Color the left-side axis title (Redemption Count) to match the pink line
    axis.title.y.left = element_text(color = "#83A0BE", size = 12),
    # Center the title and increase font size
  )

@@

@@

The data cleaning process involved filtering out some of the columns from the raw dataset and renaming some of the data entries for clarity and simplicity.

>>>>

The dataset used in this analysis is the 2024 installment of “Toronto Island Ferry Ticket Counts” obtained from OpenDataToronto (@R-opendatatoronto). This dataset contains information about the number of ferry tickets redeemed and sold for the Toronto Island Ferry system, which serves as a critical transportation link between downtown Toronto and the Toronto Islands.

@@

The data was processed and analyzed using the R programming language (@citeR). The tidyverse (@citeTidyverse) and opendatatoronto packages were employed to download, clean, and prepare the dataset for analysis. The resulting cleaned dataset was then utilized to generate the visualizations and statistical analyses presented in this paper using a collection of libraries which included the following packages:

- ggplot2 [@citeGgplot2]
- knitr [@citeKnitr]
- readr [@citeReadr]
- gridExtra, solely for the propose of plot arrangements [@R-gridExtra]
- lubridate [@R-lubridate]
- here, helps to create relative file paths [@citeHere]

@@

The dataset used in this analysis provides timestamped records of ticket redemptions and sales for the Toronto Island Ferry system, allowing for a detailed time series analysis of ferry usage patterns. @fig-linePlot1 illustrates the peak day of ticket sales for each year. Understanding these peak days and their contributing factors can help in forecasting future demand and optimizing resource allocation for similar high-traffic periods.

```{r}
#| label: fig-linePlot1
#| fig-cap: Peak Day for Each Year
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
#| fig-pos: "H"  #fix the plot in current place when render
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)

@@

@fig-linePlot1 presents the daily peak ticket sales counts for the Toronto Island Ferry system across various years, offering a clear visualization of seasonal fluctuations in ferry usage. Each line represents the total ticket sales for the busiest day of each month throughout the year, with distinct colors assigned to different years for easy comparison. Notably, the graph highlights consistent peaks in ticket sales during the summer months, especially from June to August, reflecting heightened ferry usage during the tourist season. The sharp decline in sales in 2020, evident in the orange line, corresponds to the COVID-19 pandemic's impact on public transportation and tourism activities (@urbanyi2020maritime). As restrictions eased in 2021, a gradual recovery in ticket sales is observed, depicted by the brown line, indicating a resurgence in ferry demand as tourism began to pick up in Toronto (@travel_gov_2024). This visualization not only showcases the annual variations in peak demand but also underscores the need for adaptive operational strategies to accommodate seasonal surges and unexpected disruptions.

@fig-linePlot2 on the following page shows the Redemption and Sales Counts in months over years. The graph uses dashed lines to represent the monthly sales counts and straight lines for the redemption counts. Notably, the inverted redemption counts are plotted for better visualization alongside the sales data. This figure highlights consistent seasonal patterns, with both ticket sales and redemptions peaking during the summer months, from June to August, across most years. The dip observed in 2020 reflects the significant impact of the COVID-19 pandemic on ferry operations, with reduced ferry usage compared to previous years. These trends show the importance of flexible operational planning to accommodate both predictable seasonal demand and unexpected disruptions.

```{r}
#| label: fig-linePlot2
#| fig-cap: Redemption and Sales Counts Over Time in months
#| warnings: false
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(lubridate)

@@

@fig-linePlot3 below compares the sales count and redemption count in year 2021 and 2023. The two side-by-side plots illustrate distinct usage patterns across these years.

```{r}
#| label: fig-linePlot3
#| fig-cap: Redemption and Sales Counts Over Time in 2021 and 2023
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"  # Fix the plot in current place when rendered
#| fig-width: 8
#| fig-height: 3

@@

# Plot for 2023 (Hide negative values)
plot_2023 <- ggplot(data_2023, aes(x = Timestamp)) +
  geom_line(aes(y = Redemption.Count), color = "#A977A6", linewidth = 1.2) +
  geom_line(aes(y = -Sales.Count), color = "#C1A9D2", linewidth = 1.2) +
  scale_y_continuous(
    limits = c(-y_max, y_max), 
    breaks = seq(0, y_max, by = 1000),  # Only show positive labels for Redemption Count
    sec.axis = sec_axis(~ -., name = "Sales Count", breaks = seq(0, y_max, by = 1000))  # Only show positive labels for Sales Count
  ) +
  scale_x_datetime(date_breaks = "4 months", date_labels = "%b") +  # Show only 3 labels
  labs(y = "Redemption Count", x = "Month", title = "2023") +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "#C1A9D2"),
    axis.title.y.left = element_text(color = "#A977A6")
  )

# Plot for 2021 (Hide negative values)
plot_2021 <- ggplot(data_2021, aes(x = Timestamp)) +
  geom_line(aes(y = Redemption.Count), color = "#F49600", linewidth = 1.2) +
  geom_line(aes(y = -Sales.Count), color = "#F9CB80", linewidth = 1.2) +
  scale_y_continuous(
    limits = c(-y_max, y_max), 
    breaks = seq(0, y_max, by = 1000),  # Only show positive labels for Redemption Count
    sec.axis = sec_axis(~ -., name = "Sales Count", breaks = seq(0, y_max, by = 1000))  # Only show positive labels for Sales Count
  ) +
  scale_x_datetime(date_breaks = "4 months", date_labels = "%b") +  # Show only 3 labels
  labs(y = "Redemption Count", x = "Month", title = "2021") +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "#F9CB80"),
    axis.title.y.left = element_text(color = "#F49600")
  )

# Corrected function call
grid.arrange(plot_2021, plot_2023, ncol = 2)

In 2021, ticket sales and redemptions were more variable, reflecting the gradual recovery from the COVID-19 pandemic as restrictions were lifted and ferry services resumed normal operations. Peaks and troughs are more pronounced, indicating fluctuating demand as public confidence in using public transportation returned. Conversely, the patterns in 2023 exhibit a more consistent trend, with a steadier increase in both ticket sales and redemptions during the summer months, reflecting a return to pre-pandemic demand levels. This comparison highlights the evolving dynamics in ferry usage, influenced by external factors such as public health measures and tourism trends.

As shown in @fig-linePlot4, the x-axis denotes the timestamps of each record, while the y-axis represents the number of tickets processed. The red line corresponds to the redemption counts, and the blue line represents the sales counts, plotted inversely to emphasize the contrast between the two trends. The symmetrical y-axis allows for a clearer comparison between peaks and troughs in both counts over time.

```{r}
#| label: fig-linePlot4
#| fig-cap: Redemption and Sales Counts Over Time in years
#| fig-pos: "H"
#| fig-width: 6
#| fig-height: 2.5
#| echo: false
#| warning: false
#| message: false

@@

# Create the plot
ggplot(data, aes(x = Timestamp)) +
  # Line plot for Redemption Count
  geom_line(aes(y = Redemption.Count), color = "brown2", linewidth = 1.2) +
  # Line plot for Sales Count (inverted)
  geom_line(aes(y = -Sales.Count), color = "#377EB8", linewidth = 1.2) +

  # Set y-axis limits symmetrically and apply the secondary axis
  scale_y_continuous(
    limits = c(-y_max, y_max),
    breaks = seq(0, y_max, by = 1000),  # Show only positive values for Redemption Count
    sec.axis = sec_axis(~ -., 
                        name = "Sales Count",
                        breaks = seq(0, y_max, by = 1000))  # Show only positive values for Sales Count
  ) +

@@

  # Apply minimal theme and customize axis titles
  theme_minimal() +
  theme(
    # Color the right-side axis title (Sales Count) to match the Sales line
    axis.title.y.left = element_text(color = "brown3", size = 12),
    # Color the left-side axis title (Redemption Count) to match the Redemption line
    axis.title.y.right = element_text(color = "dodgerblue3", size = 12),
    # Center the title and increase font size
  )

@@

While the analysis provides valuable insights, there are some limitations that should be acknowledged. The dataset does not account for external variables such as weather conditions or special events, which may also significantly impact ferry usage. Additionally, the data is limited to ticket counts and does not capture other relevant factors such as passenger demographics or satisfaction levels.

@@

The raw FerrySale dataset underwent several cleaning steps to ensure its accuracy and usability for analysis. Initially, all rows with missing values were removed, followed by converting the Timestamp column to a standardized date-time format for consistent time series analysis. Duplicate entries were identified and eliminated to prevent inflated ticket counts. Additionally, transactions with unrealistic values, such as negative counts for ticket redemptions or sales, were filtered out. The resulting cleaned dataset, saved as cleaned_data.csv, provided a reliable foundation for the statistical analyses and visualizations presented in this paper.
